---
title: "How To Advance Science Using _Bioconductor_"
author:
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, New York
  email: Martin.Morgan@RoswellPark.org
package: BiocIntro
output:
  BiocStyle::html_document
abstract: |
  Now is an incredibly exciting time for you to participate in
  bioinformatic research, with new methods for generating and
  analyzing large genomic data sets emerging almost daily. As
  bioinformaticians, we need a large, comprehensive, current,
  established, well supported, open source, community developed
  collection of software to address our leading-edge needs. This
  presentation describes the _Bioconductor_ project, providing exactly
  these resources for the _R_ bioinformatics community. We'll learn a
  little about the history, philosophy, and approach of the
  _Bioconductor_ project. We then walk through essential steps in
  getting started with _Bioconductor_, tackling real analytic
  challenges, and contributing to the _Bioconductor_
  community. With luck, the presentation will inspire and empower you
  to tackle new and innovative challenges in your own bioinformatic
  research.
vignette: |
  %\VignetteIndexEntry{How To Advance Science Using Bioconductor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
    cache = as.logical(Sys.getenv("KNITR_CACHE", TRUE)),
    collapse = TRUE
)
```

# _Bioconductor_

## Overview

Statistical analysis and comprehension of high-throughput genomic data

- https://bioconductor.org
- https://support.bioconductor.org

Open source, open development

- Core team supported by US NIH & other funding

Origin, participation, and maintenance

- 1823 software packages (approximately 100 new packages every 6 months)
- 1/2 million unique IP downloads annually; 33k PubMedCentral citations
- Global community

## Scientific domains

Historical development: core

- Microarrays, (bulk) RNA-sequencing, single-cell genomics

Domains

- Differential expression, epigenomics, called variants, flow
  cytometry, proteomics, metabolomics
  
## Unique features

Vignettes & reproducible research

Interoperability

- Structured data representation via 'S4' classes

Annotation and experiment data resources

# Getting science done

## Getting started

_R_ & [BiocManager][]

## Tackling real problems

Packages, vignettes, and work flows

[OSCA][]: Orchestrating single-cell analysis in _R_ / _Bioconductor_

## Contributing to the community

Package contributions & the build system

Dissemination and support

Developer resources

- [community-bioc][] slack ([sign-up][])

# Opportunities

Domain-specific

Container and cloud computing

- '[Bioconductor full][]' and other Docker images
- The [AnVIL][]: cloud computing adjacent to consortium-scale data

# Acknowledgements

Research reported in this presentation was supported by the NCI and
NHGRI of the National Institutes of Health under award numbers
U24CA232979, U41HG004059, and U24CA180996.  The content is solely the
responsibility of the authors and does not necessarily represent the
official views of the National Institutes of Health.

```{r sessionInfo}
sessionInfo()
```

[BiocManager]: https://cran.r-project.org/package=BiocManager
[OSCA]: https://osca.bioconductor.org
[community-bioc]: https://community-bioc.slack.com
[sign-up]: https://bioc-community.herokuapp.com/
[AnVIL]: https://anvil.terra.org
[Bioconductor full]: https://github.com/Bioconductor/bioconductor_full
