---
title: "Bioconductor Package Development"
author:
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: BiocIntro
vignette: |
  %\VignetteIndexEntry{A01 -- Bioconductor Package Development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

# Packages

## What and Why?

What?

- A simple directory structure with text files.
- `DESCRIPTION`: title, author, version, license, etc.
- `NAMESPACE`: functions used by and made available by your package
- `R/`: function defintions
- `man/`: help pages
- `vignettes/`: vignettes
- `tests/`: code to test your package

Why?

- Organize an analysis.
- Share reproducible code with lab mates, colleagues, ...

Minimal

```
$ tree MyPackage
MyPackage
└── DESCRIPTION

0 directories, 1 file
$ cat MyPackage/DESCRIPTION 
Package: MyPackage
Type: Package
Version: 0.0.1
Author: Martin Morgan
Maintainer: Martin Morgan <martin.morgan@roswellpark.org>
Title: A Minimal Package
Description: An abstract-like description of the package.
License: Artistic-2.0
```

Typical

```
$ tree
.
└── MyPackage
    ├── DESCRIPTION
    ├── man
    │   └── hi.Rd
    ├── NAMESPACE
    ├── R
    │   └── hi.R
    ├── tests
    │   ├── testthat
    │   │   └── test_hi.R
    │   └── testthat.R
    └── vignettes
        └── MyPackage.Rmd

6 directories, 7 files
```

## Working with packages

Build


```
$ R CMD build MyPackage
* checking for file 'MyPackage/DESCRIPTION' ... OK
* preparing 'MyPackage':
* checking DESCRIPTION meta-information ... OK
* checking for LF line-endings in source and make files and shell scripts
* checking for empty or unneeded directories
* creating default NAMESPACE file
* building 'MyPackage_0.0.1.tar.gz'
```

Check

```
$ R CMD check MyPackage_0.0.1.tar.gz 
* using log directory '/home/mtmorgan/a/BiocIntro/vignettes/MyPackage.Rcheck'
* using R version 3.4.2 Patched (2017-10-12 r73550)
* using platform: x86_64-pc-linux-gnu (64-bit)
* using session charset: UTF-8
* checking for file 'MyPackage/DESCRIPTION' ... OK
* checking extension type ... Package
* this is package 'MyPackage' version '0.0.1'
* checking package namespace information ... OK
* checking package dependencies ... OK
* checking if this is a source package ... OK
* checking if there is a namespace ... OK
* checking for executable files ... OK
* checking for hidden files and directories ... OK
* checking for portable file names ... OK
* checking for sufficient/correct file permissions ... OK
* checking whether package 'MyPackage' can be installed ... OK
* checking installed package size ... OK
* checking package directory ... OK
* checking DESCRIPTION meta-information ... OK
* checking top-level files ... OK
* checking for left-over files ... OK
* checking index information ... OK
* checking package subdirectories ... OK
* checking whether the package can be loaded ... OK
* checking whether the package can be loaded with stated dependencies ... OK
* checking whether the package can be unloaded cleanly ... OK
* checking whether the namespace can be loaded with stated dependencies ... OK
* checking whether the namespace can be unloaded cleanly ... OK
* checking loading without being on the library search path ... OK
* checking examples ... NONE
* checking PDF version of manual ... OK
* DONE

Status: OK
```

Install

```
$ R CMD INSTALL MyPackage_0.0.1.tar.gz 
* installing to library '/home/mtmorgan/R/x86_64-pc-linux-gnu-library/3.4-Bioc-3.6'
* installing *source* package 'MyPackage' ...
** help
No man pages found in package  'MyPackage' 
*** installing help indices
** building package indices
** testing if installed package can be loaded
* DONE (MyPackage)
```

# Package Development

## Olde School

- Add new functions in files `R/foo.R`
- Update `NAMESPACE` to import functions or packages used by your
  function, and to export your functions that users will want to use.
- Create `man` pages by hand
- Write vignettes in LaTeX Sweave
- Key reference: _Writing R Extensions_, `RShowDoc("R-exts")`

## New School

- `devtools::create()` a package skeleton. More flexible `Authors@R`
  instead of `Author:` / `Maintainer:` fields.
- Use 'roxygen' to document functions

    - Lines starting with `#'` are documentation lines
    - `@details`, `@param`, `@return`, `@examples` document the function
    - `@export` indicates that the function should be visible to the user
    - `@import` and `@importFrom` indicate (non-base) functions that
      are used by this function, e.g,. `@importFrom stats rnorm runif`
    - `devtools::document()` to update documentation.

    ```
    #' Title, e.g., Say 'hi' to friends.
    #'
    #' Short description of this help page. `hi("Martin")` returns a greeting.
    #'
    #' @details A more extensive description of the functions or other objects
    #'    documented on this help page. Use `how=` to determine the nature of
    #'    the greeting.
    #'
    #' @param who character() The name(s) of the person / people to greet.
    #'
    #' @param how character(1) Whether to shout (uppercase) or whisper 
    #'     (lowercase) the greeting.
    #'
    #' @return character() of greetings, with length equal to `who`.
    #'
    #' @examples
    #' hi(c("Martin", "Jenny"), "whisper")
    #'
    #' @export
    hi <- function(who, how = c("asis", "shout", "whisper")) {
        stopifnot(
            is.character(who),
            is.character(how), missing(how) || length(how) == 1
        )
        transform <- switch(
            match.arg(how),
            asis = identity, shout = toupper, whisper = tolower
        )

        greet <- paste("hi", who)
        transform(greet)
    }
    ```
    
- Build / check / install using devtools from an _R_ session inside
  the package folder.

    ```
    getwd()  # e.g., MyPackage/
    devtools::build()
    devtools::check()
    devtools::install()
    ```
    
- During development, short-circuit full round-trip with
  `devtools::load_all()`
  
- Write vignettes in markdown, e.g.,
  `vignettes/MyPackage.Rmd`. `usethis::use_vignette()` (see also
  [BiocStyle][]).

    ```{r, echo=FALSE, comment=""}
    fl <- system.file(
        package = "BiocIntro",
        "MyPackage", "vignettes", "MyPackage.Rmd"
    )
    cat(readLines(fl), sep="\n")
    ```

- Write _unit tests_ that validate the correctness of you
  functions. `usethis::use_testthat()`
  
    ```
    $ tree
    .
    ...
    ├── tests
    │   ├── testthat
    │   │   └── test_hi.R
    │   └── testthat.R
    ...
    ```

    Content of `test_hi.R`:

    ```{r, echo=FALSE, comment=""}
    fl <- system.file(
        package = "BiocIntro",
        "MyPackage", "tests", "testthat", "test_hi.R"
    )
    cat(readLines(fl), sep="\n")
    ```
  
[BiocStyle]: https:/bioconductor.org/packages/BiocStyle

# Contributing to _Bioconductor_

Process

- Open an issue at https://github.com/Bioconductor/Contributions
- Automatic review -- `R CMD check`, `R CMD BiocCheck` (after
  `biocLite("BiocCheck")`)
- Manual review

Manual review

- Open process
- Basic code review

Common comments

- Re-use, e.g., `rtracklayer::import.bed()`, rather than re-invent.
- Inter-operate, e.g,. use `SummarizedExperiment()` rather than `matrix`.
- Use portable code, e.g,. `tempfile()` rather than hard-coded path
- Use robust code, e.g., `seq_len(n)` rather than `1:n`
- 'Vectorize' instead of iterate, e.g,. `sqrt(x)` instead of `sapply(x, sqrt)`.
- Avoid high 'cyclomatic complexity'

    - Only one or a few paths through a function.
    - Assertions about inputs at the start of the function, not
      part-way through.
    - Choose and write functions that are vectorized, and that handle
      the edge cases, e.g., length 0 arguments or `NA` values,
      correctly (compare `sapply(integer(), sqrt)` with
      `vapply(integer(), sqrt)`).

- Functions 'fit in your head', literally.

    - Refactor to allow re-use rather than repetition of common code.
    - Refactor to isolate logically consistent operations -- testable
      inputs and outputs.

# Best Practices

## `man` pages

- Document functions on man pages that have the same names as
  corresponding `.R` files.
- Specify parameters (input arguments) using standard _R_ idioms
  (e.g., `character(1)`).
- Specify return value.
- Easy-to-run, illustrative examples.

## Vignettes

- Overall use and interoperation with other packages / stages in the
  work flow.
- 'Toy' data for easy reproducibility, but realistic enough to
  illustrate nuannces.

## Unit tests

- Use [testthat][] or other packages.
- Avoid using examples or vignette to test edge-cases; this just
  confuses the user.

## Version control

- Use git for local version control; consider [github][] for sharing.

[testthat]: https://cran.r-project.org/?pacakge=testthat
[github]: https://github.com
