---
title: "ChIP-seq Analysis"
author: Martin Morgan <martin.morgan@roswellpark.org>
date: "18 September, 2018"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{ChIP-seq Analysis}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE"))
)
```

```{r setup, echo=FALSE}
suppressPackageStartupMessages({
    library(ChIPseeker)
    library(TxDb.Hsapiens.UCSC.hg19.knownGene)
})
```

# Introduction

Based on the [ChIPseeker][] 'pipeline' vignette.

[ChIPseeker]: https://bioconductor.org/packages/ChIPseeker

# Working with called peaks -- 'bed' files

```{r}
library(ChIPseeker)
fls <- dir("~/extdata", "*bed.gz", full = TRUE)
peak <- readPeakFile(fls[[4]])
peak
covplot(peak)
```

# Peaks in promoters

```{r}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
promoter <- getPromoters(TxDb=txdb, upstream=3000, downstream=3000)
tagMatrix <- getTagMatrix(peak, windows=promoter)
dim(tagMatrix)
plot(colSums(tagMatrix), type="l")
```

```{r}
tagHeatmap(tagMatrix, xlim=c(-3000, 3000))
```

# Annotating peaks

```{r}
peakAnno <- annotatePeak(
    fls[[4]], tssRegion=c(-3000, 3000), TxDb=txdb, annoDb="org.Hs.eg.db"
)
```

```{r}
plotAnnoPie(peakAnno)
```

```{r}
upsetplot(peakAnno)
```

# Differential binding

# Session info

As a part of the effort to provide reproducible analysis pipelines, it
is essential to document the software and versions used; here is the
information about _R_ packages used in the current chapter.

```{r}
sessionInfo()
```
