---
title: "Advanced and Emerging Topics"
author: Martin Morgan <martin.morgan@roswellpark.org>
date: "18 September, 2018"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{Advanced and Emerging Topics}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE"))
)
```

```{r setup, echo=FALSE}
suppressPackageStartupMessages({
    library(SummarizedExperiment)
    library(TENxBrainData)
    library(BiocParallel)
})
```

# Single Cell Analysis

- Check out the [simpleSingleCell][] [workflow][]

[simpleSingleCell]: https://bioconductor.org/packages/simpleSingleCell
[workflow]: https://bioconductor.org/packages

# Very Large Data

## Parallel evaluation

```{r}
library(BiocParallel)
system.time(
    lapply(1:4, function(i) Sys.sleep(1))
)
system.time(
    bplapply(1:4, function(i) Sys.sleep(1))
)
```

Cores or clusters?

- Cores: single computer

    ```{r}
    param <- MulticoreParam(workers = 4)
    results <- bplapply(1:4, function(i) Sys.sleep(1), BPPARAM = param)
    ```

- Clusters -- job scheduler like _slurm_

    ```{r, eval = FALSE}
    param <- BatchtoolsParam(workers = 4, cluster="slurm")
    results <- bplapply(1:4, function(i) Sys.sleep(1), BPPARAM = param)
    ```

## 'on-disk' data

```{r}
library(TENxBrainData)
suppressMessages({
    tenx <- TENxBrainData()
})
tenx
assay(tenx)
hist(log10(colSums(assay(tenx)[,1:1000])))
```

# Session info

As a part of the effort to provide reproducible analysis pipelines, it
is essential to document the software and versions used; here is the
information about _R_ packages used in the current chapter.

```{r}
sessionInfo()
```
