---
title: "A.2 -- Data Input and Manipulation"
author: Martin Morgan <martin.morgan@roswellpark.org>
date: "21 - 22 January, 2016"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{A.2 -- Data Input and Manipulation}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
options(width=100)
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

# Extended Exercise 1: BRFSS Survey Data

We will explore a subset of data collected by the CDC through its
extensive Behavioral Risk Factor Surveillance System ([BRFSS][])
telephone survey. Check out the link for more information. We'll look
at a subset of the data.

1. Use `file.choose()` to find the path to the file 'BRFSS-subset.csv'

    ```{r file.choose, eval=FALSE}
    path <- file.choose()
    ```
    
    ```{r system.file, echo=FALSE}
    path <- system.file(package="BiocIntroRPCI", "extdata", "BRFSS-subset.csv")
    ```

2. Input the data using `read.csv()`, assigning to a variable `brfss`

    ```{r read.csv}
    brfss <- read.csv(path)
    ```

3. Use command like `class()`, `head()`, `dim()`, `colnames()`,
   `summary()` to explore the data.

    - What variables have been measured?

    - Can you guess at the units used for, e.g., Weight and Height?

4. Use the `$` operator to extract the 'Sex' column, and summarize the
   number of males and females in the survey using `table()`. Do the
   same for 'Year'.

    ```{r brfss-sex}
    table(brfss$Sex)
    ```

5. The `xtabs()` function performs cross-tabulation using a
   formula-like interface; summarize the number of males and female in
   each year of the study.

    ```{r brfss-xtabs}
    xtabs(~ Year + Sex, brfss)
    ```

6. Use `aggregate()` to summarize the mean weight of each group. What
   about the median weight of each group?

    ```{r brfss-aggregate}
    aggregate(Weight ~ Year + Sex, brfss, mean)
    ```

7. Create a subset of the data consisting of only the 1990
   observations. Perform a t-test comparing the weight of males and
   females ("'Weight' as a function of 'Sex'", `Weight ~ Sex`)

    ```{r t-test-1990}
    brfss_1990 = brfss[brfss$Year == 1990,]
    t.test(Weight ~ Sex, brfss_1990)
    ```
   What about differences between weights of males (or females) in
   1990 versus 2010? Check out the help page `?t.test.formula`. Is
   there a way of performing a t-test on `brfss` without explicitly
   creating the object `brfss_1990`?

8. Use `boxplot()` to plot a histogram of weights of the 1990 Female
   individuals. Can you transform weight, e.g., `sqrt(Weight) ~ Year`?
   Interpret the results. Do similar boxplots for the t-tests of the
   previous question.

    ```{r brfss-boxplot, fig.width=5, fig.height=5}
    boxplot(Weight ~ Year, brfss, subset = (Sex == "Male"),
            main="Males")
    ```

[BRFSS]: http://www.cdc.gov/brfss/about/index.htm

# Extended Exercise 2: ALL Phenotypic Data

This data comes from an (old) Acute Lymphoid Leukemia‎ microarray data
set.
