---
title: "A.3 -- Statistics and Graphics"
author:
  Martin Morgan <Martin.Morgan@RoswellPark.org><br/>
date: "11 - 12 September 2017"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{A.3 -- Statistics and Graphics}
  % \VignetteEngine{knitr::rmarkdown}
---

```{r style, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE"))
)
suppressPackageStartupMessages({
    library(tidyverse)
})
```

# Exploration and simple univariate measures

Input clean data, with `Sex` and `Year` as factors.

<!--
```{r echo=FALSE}
path <- file.path("extdata", "BRFSS-subset.csv")
```
-->
```{r ALL-choose, eval=FALSE}
path <- file.choose()    # look for BRFSS-subset.csv
```

```{r ALL-input}
stopifnot(file.exists(path))
library(tidyverse)
col_types <- cols(
    Age = col_integer(),
    Weight = col_double(),
    Sex = col_factor(c("Female", "Male")),
    Height = col_double(),
    Year = col_factor(c("1990", "2010"))
)
brfss <- read_csv(path, col_types = col_types)
brfss
```

## Univariate: `t.test()` for Weight in 1990 vs. 2010 Females

Filter the data to include females only, and use base `plot()` function
and the formula interface to visualize the relationship between
`Weight` and `Year`.

```{r brfss-female-plot}
brfss %>% filter(Sex %in% "Female") %>% plot(Weight ~ Year, data = .)
```

Use a `t.test()` to test the hypothesis that female weight is the same
in 2010 as in 1990

```{r brfss-female-t-test}
brfss %>% filter(Sex %in% "Female") %>% t.test(Weight ~ Year, data = .)
```

## Bivariate: Weight and height in 2010 Males

Filter the data to contain 2010 Males. Use `plot()` to visualize the
relationship, and `lm()` to model it.

```{r brfss-male}
male2010 <- brfss %>% filter(Year %in% "2010", Sex %in% "Male")
male2010 %>% plot( Weight ~ Height, data = .)
fit <- male2010 %>% lm( Weight ~ Height, data = .)
fit
summary(fit)
```

Multiple regression: Weight and Height, accounting for difference between years

```{r brfss-male-year-and-height}
male <- brfss %>% filter(Sex %in% "Male")
male %>% lm(Weight ~ Year + Height, data = .) %>% summary()
```

Is there an interaction between `Year` and `Height`?


```{r brfss-male-interaction}
male %>% lm(Weight ~ Year * Height, data = .) %>% summary()
```

Check out other things to do with fitted model:

- `broom::tidy()`: P-value, etc., as data.frame
- `broom::augment()`: fitted values, residuals, etc

    ```{r brfss-male-augment, warning=FALSE}
    library(broom)
    male %>% lm(Weight ~ Year + Height, data = .) %>% 
        augment() %>% as.tibble()
    ```

## Visualization: [ggplot2][]

*gg*plot: "Grammar of Graphics"

- data: `ggplot2()`
- *aes*thetics: `aes()`, 'x' and 'y' values, point colors, etc.
- *geom*metric summaries, layered
    - `geom_point()`: points
    - `geom_smooth()`: fitted line
    - `geom_*`: ...
- *facet* plots (e.g., `facet_grid()`) to create 'panels' based on
  factor levels, with shared axes.

Create a plot with data points

```{r male-geom_point, warning = FALSE}
ggplot(male, aes(x=Height, y = Weight)) + geom_point()
```

Capture the base plot and points, and explore different smoothed
relationships, e.g., linear model, non-parameteric smoother

```{r male-ggplot, warning = FALSE}
plt <- ggplot(male, aes(x=Height, y = Weight)) + geom_point()
plt + geom_smooth(method = "lm")
plt + geom_smooth()       # default: generalized additive model
```

Use an `aes()`thetic to color smoothed lines based on `Year`, or
`facet_grid()` to separate years.

```{r male-facet, warning = FALSE}
plt + geom_smooth(method = "lm", aes(color = Year))
plt + facet_grid( ~ Year ) + geom_smooth(method = "lm")
```

[ggplot2]: https://cran.r-project.org/package=ggplot2

# Multivariate analysis

This is a classic microarray experiment. Microarrays
consist of 'probesets' that interogate genes for their level of
expression. In the experiment we're looking at, there are 12625
probesets measured on each of the 128 samples. The raw expression
levels estimated by microarray assays require considerable
pre-processing, the data we'll work with has been pre-processed.

## Input and setup

Start by finding the expression data file on disk.

<!--
```{r echo=FALSE}
path <- file.path("extdata", "ALL-expression.csv")
stopifnot(file.exists(path))
```
-->

```{r ALL-choose-again, eval=FALSE}
path <- file.choose()          # look for ALL-expression.csv
stopifnot(file.exists(path))
```

The data is stored in 'comma-separate value' format, with each
probeset occupying a line, and the expression value for each sample in
that probeset separated by a comma. Input the data using
`read.csv()`. There are three challenges:

1. The row names are present in the first column of the data. Tell _R_
   this by adding the argument `row.names=1` to `read.csv()`.
2. By default, _R_ checks that column names do not look like numbers,
   but our column names _do_ look like numbers. Use the argument
   `check.colnames=FALSE` to over-ride _R_'s default.
3. `read.csv()` returns a `data.frame`. We could use a `data.frame` to
   work with our data, but really it is a `matrix()` -- the columns
   are of the same type and measure the same thing. Use `as.matrix()`
   to coerce the `data.frame` we input to a `matrix`.

```{r ALL-input-exprs}
exprs <- read.csv(path, row.names=1, check.names=FALSE)
exprs <- as.matrix(exprs)
class(exprs)
dim(exprs)
exprs[1:6, 1:10]
range(exprs)
```

We'll make use of the data describing the samples

<!--
```{r echo=FALSE}
path <- file.path("extdata", "ALL-phenoData.csv")
stopifnot(file.exists(path))
```
-->

```{r ALL-phenoData.csv-clustering-student, eval=FALSE}
path <- file.choose()         # look for ALL-phenoData.csv
stopifnot(file.exists(path))
```

```{r}
pdata <- read.csv(path, row.names=1)
class(pdata)
dim(pdata)
head(pdata)
```

Some of the results below involve plots, and it's convenient to choose
pretty and functional colors. We use the [RColorBrewer][]
package; see [colorbrewer.org][]

[RColorBrewer]: https://cran.r-project.org/?package=RColorBrewer
[colorbrewer.org]: http://colorbrewer.org

```{r colors}
library(RColorBrewer)  ## not available? install package via RStudio
highlight <- brewer.pal(3, "Set2")[1:2]
```

`highlight' is a vector of length 2, light and dark green.

For more options see `?RColorBrewer` and to view the predefined
palettes `display.brewer.all()`

## Cleaning

We'll add a column to `pdata`, derived from the `BT` column, to
indicate whether the sample is B-cell or T-cell ALL.

```{r ALL-BorT}
pdata$BorT <- factor(substr(pdata$BT, 1, 1))
```

Microarray expression data is usually represented as a matrix of genes
as rows and samples as columns. Statisticians usually think of their
data as samples as rows, features as columns. So we'll transpose the
expression values

```{r}
exprs <- t(exprs)
```

Confirm that the `pdata` rows correspond to the `exprs` rows.

```{r}
stopifnot(identical(rownames(pdata), rownames(exprs)))
```

## Unsupervised machine learning -- multi-dimensional scaling

Reduce high-dimensional data to lower dimension for visualization.

Calculate distance between _samples_ (requires that the expression
matrix be transposed).

```{r}
d <- dist(exprs)
```

Use the `cmdscale()` function to summarize the distance matrix into
two points in two dimensions.

```{r}
cmd <- cmdscale(d)
```

Visualize the result, coloring points by B- or T-cell status

```{r}
plot(cmd, col=highlight[pdata$BorT])
```
