---
title: "Exploring Data"
author:
- name: Martin Morgan
  affiliation: Roswell Park Cancer Institute, Buffalo, NY
output:
  BiocStyle::html_document:
    toc_float: true
package: BiocStyle
vignette: |
  %\VignetteIndexEntry{Working with Tidy Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(tidyverse)
library(magrittr)
```

# Univariate

- `summary()`
- `fivenum()`
- `t.test()`
- `chisq.test()`
- `xtabs()`

## Survival analysis

Data input

```{r}
library(tidyverse)

col_types <- cols(
    SAMPLE_TYPE = col_factor(c("Metastatic", "Primary Tumor")),
    OS_STATUS = col_factor(c("DECEASED", "LIVING")),
    DFS_STATUS = col_factor(c("DiseaseFree", "Recurred/Progressed"))
)
clean <- read_csv(
    "extdata/skcm_tcga_rna_seq_v2_mrna_median_Zscores.csv",
    col_types = col_types
)
```

Model fit

```{r}
library(survival)
library(broom)   # tidy()
surv <- clean %>% with(Surv(OS_MONTHS, OS_STATUS == "DECEASED"))
fits <- bind_rows(
    clean %>% coxph(surv ~ CD63, data = .) %>% tidy(),
    clean %>% coxph(surv ~ CD9, data = .)  %>% tidy(),
    clean %>% coxph(surv ~ CD81, data = .) %>% tidy()
) %>% as_tibble()
fits
```

Visualization as Kaplan-Meier curves

- Treat gene expression as _categorical_ instead of continuous

```{r}
library(survminer)

genes <- c("CD63", "CD9", "CD81")
cutpoints <- clean %>% 
    mutate(OS_STATUS = OS_STATUS == "DECEASED") %>% 
    surv_cutpoint("OS_MONTHS", "OS_STATUS", genes)
cats <- cutpoints %>% surv_categorize() %>% as_tibble()

surv <- cats %>% with(Surv(OS_MONTHS, OS_STATUS))
fit <- survfit(surv ~ CD81, data = cats)
ggsurvplot(fit, conf.int=TRUE)
```

# Multivariate

- [Multivariate][] task view
- `lm()` / `summary()` / `anova()` / `broom::tidy()`
- `cmdscale()` -- multidimensional scaling / principal coordinates analysis
- clustering: `hclust()`, `kmeans()`, ... [Clustering][] task view
- classification: `class::knn()`, ... [Machine Learning][] task view

[Multivariate]: https://cran.r-project.org/view=Multivariate
[Clustering]: https://cran.r-project.org/view=Clustering
[Machine Learning]: https://cran.r-project.org/view=MachineLearning
