---
title: "Programming Concepts"
author:
- name: Martin Morgan
  affiliation: Roswell Park Cancer Institute, Buffalo, NY
output:
  BiocStyle::html_document:
    toc_float: true
package: RIntro
vignette: |
  %\VignetteIndexEntry{Programming Concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Case study: MDS on gene expression data

Data input

```{r}
suppressPackageStartupMessages({
    library(tidyverse)
    library(magrittr)
})

pdata <- read_csv("extdata/ALL-phenoData.csv")
pdata %<>% 
    filter(mol.biol %in% c("BCR/ABL", "NEG")) %>%
    mutate(
        BT = factor(substr(BT, 1, 1)),
        mol.biol = factor(mol.biol)
    )

exprs <- read_csv("extdata/ALL-expression.csv")
exprs %<>% gather("Sample", "Expression", -Gene)
```

Script to perform MDS, keeping track of samples -- tedious and
error-prone, but frequently useful

```{r}
## from tibble to classic data structures...
classic <- exprs %>% spread("Gene", "Expression")
df <- classic %>% select(-Sample) %>% as.data.frame()
rownames(df) <- classic$Sample

## MDS
cmd <- cmdscale(dist(df))

## ... and back to tibble
cmd %<>% as.data.frame() %>% rownames_to_column("Sample") %>% as_tibble()
```

```{r}
mdstbl <- 
    function(data, id, key, value)
{
    ## from tibble to classic data structures...
    classic <- exprs %>% spread(key, value)
    df <- classic %>% select(-starts_with(id)) %>% as.data.frame()
    rownames(df) <- classic[[id]]

    ## MDS
    cmd <- cmdscale(dist(as.matrix(df)))

    ## ... and back to tibble
    cmd %<>% as.data.frame() %>% rownames_to_column(id) %>% as_tibble()
    
    cmd
}
```

# Elements of programming

functions 

```{r}
fun <- function(x, y = 1, ...) {}
```

- arguments

    - required
    - default values
    - additional

- 'first class' objects in _R_

iteration

- `apply()`, `lapply()`, `sapply()`, `vapply()`
- `for () {}`
- `while () {}`, `repeat {}`
- iteration versus vectorization -- `n` function calls versus 1
  function call.

conditional

- `if () {} else {}` (versus `ifelse()`)

# Correctness and performance

- Most important to write _correct_ code; compare to expectation with
  `identical()`, `all.equal()`.
- Measure speed with `system.time({})` or [microbenchmark][].

[microbenchmark]: https://cran.r-project.org/package=microbenchmark
