---
title: "_R_ Packages for Communicating Reproducible Research"
author:
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{R Packages for Communicating Reproducible Research}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: true
---

Last modified: 31 May, 2019

```{r style, echo = FALSE, results = 'asis'}
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE"))
)
options(width = 75)
```

# Abstract

This tutorial is for all who wish to write _R_ packages. _R_ is a fantastic language for you to develop new statistical approaches for the analysis and comprehension of real-world data. _R_ packages provide a way to capture your new approach in a reproducible, documented unit. An _R_ package is surprisingly easy to create, and creating an _R_ package has many benefits. In this tutorial we create an _R_ package. We start with a data set and a simple script transforming the data in a useful way; perhaps you have your own data set and script? We replace the script with a _function_, and place the function and data into an _R_ _package_. We then add documentation, so that our users (and our future selves) understand what the function does and how the function applies to new data sets. With an _R_ package in hand, we can tackle more advance challenges: _vignettes_ for rich narrative description of the package; _unit tests_ to make our package more robust; and _version control_ to document how we change the package. The final step in the development of our package is to share it with others, through github, through CRAN, or though domain-specific channels such as _Bioconductor_.


# _R_ packages

# The basics

## Script to function

## Package skeleton

## Functions

## Data sets

## Documentation

## Build & check

# Advanced challenges

## Communicate: vignettes

## Robust: tests & examples

## Mature: version control

## Dissemination

# `sessionInfo()`

```{r}
sessionInfo()
```
