---
title: "Working with Tidy Data"
author:
- name: Martin Morgan
  affiliation: Roswell Park Cancer Institute, Buffalo, NY
output:
  BiocStyle::html_document:
    toc_float: true
package: RIntro
vignette: |
  %\VignetteIndexEntry{Working with Tidy Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Tidy data and analysis

[Tidy data][]

- Each variable a column
- Each observation a row
- Each observational unit a table

The tidyverse

- install (once only): `install.packages("tidyverse")`
- use

    ```{r}
    library(tidyverse)
    library(magrittr)
    search()
    ```

What?

- [tibble][], [readr][], [magrittr][], [dplyr][], [ggplot2][]

Why?

- tibbles

    ```{r}
    mtcars
    as_tibble(mtcars)
    ```

- input

    ```{r}
    brfss = read_csv("~/extdata/brfss.csv")
    ```

- pipes

    ```{r}
    brfss %>% colnames()
    brfss %$% Age %>% mean(na.rm=TRUE)
    ```

- plyr

    - `filter()`
    - `select()`
    - `mutate()`
    - `group_by()`
    - `arrange()`
    - Examples: average Weight, Height, Age by Sex and Year
    - `*join()`: tables

- ggplot2

    - `ggplot()`
    - `geom_*()`
    - `facet_grid()`
    - Example: boxplot of Weight by Year, facet by Sex

Principle: correct early

    - 'Sex' and 'Year' should be factors
    - `cols()` and `read_csv()`

[Tidy data]: https://www.jstatsoft.org/article/view/v059i10
[tibble]: https://cran.r-project.org/package=tibble
[dplyr]: https://cran.r-project.org/package=dplyr
[magrittr]: https://cran.r-project.org/package=magrittr
[readr]: https://cran.r-project.org/package=readr
[ggplot2]: https://cran.r-project.org/package=ggplot2
